A finite-state morphological transducer for Kyrgyz

### 2012-03-14 paper/paper.tex
All table captions to top
more fixed FIXMEs

%FIXME:
%\thanks{GSoC, Tolgonay, ..?}

@@

%FIXED: Should we say something about how I sometimes consulted Tolgonay, or about where my knowledge of Kyrgyz comes from in the first place?
%FIXME: Tolgonay's name to acknowledgements (where?) for helping with Kyrgyz
The understanding of Kyrgyz grammar employed to construct the transducer for this project was gained from the Kyrgyz knowledge of two of our authors.  Mirlan Ipasov is a native speaker of Kyrgyz, and Jonathan Washington is a theoretical and descriptive linguist fluent in Kyrgyz and knowledgeable about Turkic languages.  Some grammar sources were consulted, such as \cite{hebertpoppe1963}, \cite{usonalievomuraliev2003}, \cite{qudaybergenov1980}, \cite{somfaikara2003}, and \cite{imart1981}, but they were largely not relied on due to their approaches to Kyrgyz grammar.  Some dictionaries were also consulted, including \cite{jumakunova2005} and \citetwoyears{yudakhin1957}{yudakhin1965}.

\subsection{Morphological transducers}
%FIXME: this seems to be wrapping badly, at least here
The objective of a morphological transducer is twofold. Firstly to take surface forms (e.g., алдым) and generate all possible lexical forms, and secondly to take lexical forms (e.g.,  ал{\tt {\small <v><tv><ifi><p1><sg>}}, алд{\tt {\small <n><px1sg><nom>}}, etc.) and generate one or more surface forms.

@@

\section{Description}
%FIXME: \ldots{} section
The tagset consists of 127 separate tags, 19 covering the main parts of speech (noun, verb, adjective, adverb, postposition, etc.) and 108 covering morphological subcategorisation for e.g. case, number, person, possession, transitivity, tense-aspect-mood, etc. The tags are represented as multicharacter symbols, between less than `<' and greater than `>' symbols. The tagset is quite extensive and still not entirely stabilised, as such a full listing is not included here. However, the tags are listed in the source code of the transducer, along with comments describing their usage.

@@

%TODO: for a future version
%\subsection{Challenges:  Copula, how to represent different tenses, what is a “word”?}

@@

The first exception does not apply to Kyrgyz, as unlike other Turkic languages (Tatar, Chuvash, Kazakh), although it has morphophonological processes that work across the `whitespace' boundary, these are not represented in the orthography.

@@

%FIXME: why copular suffixes?

@@

\footnotetext{P1--P6 refer to different sets of personal suffixes; the terminology and specific numbers are based off of \cite{hebertpoppe1963}[29].}

@@

\begin{table}
	\centering
	\begin{tabular}{lr}
		\toprule
		Part of speech & Number of stems\\
		\midrule
		Noun & 4,972\\
		Verb & 1,231 \\
		Adjective & 944\\
		Proper noun & 796\\
		Adverb & 295 \\
		Numeral & 63 \\
		Conjunction & 58 \\
		Postposition & 51 \\
		Pronoun & 29\\
		Determiner & 27\\
		\midrule
		Total: & 8,466 \\
		\bottomrule
	\end{tabular}
	\caption{Breakdown of number of stems in the lexicon by major part of speech. Minor categories, such as particle, modal and copula are left out.}\label{table:stems}
	%FIXME: are we putting captions above or below tables?  I like them better below, but I got the impression that there's some standard to have them above.  Maybe it was just UW being table Nazis for my MA thesis.
\end{table}

@@

%FIXME: is this about right?  and I mean all aspects of the paragraph
%FIXME: ZZ needs to be replaced with .. ?
To calculate the naïve coverage\footnote{Naïve coverage refers to the percentage of surface forms in a given corpora that receive at least one analysis.  Forms counted by this measure may have other analyses which are not delivered by the transducer.} of the analyser, two corpora were used.  The first corpus was a database dump of the Kyrgyz Wikipedia, dated 2011-09-23,\footnote{The exact name of the dump was \texttt{kywiki-20110923-pages-articles.xml}.} which was processed with the programs \texttt{aq-wikicrp} to extract sentences. The second was a corpus generated from the archives of Radio Free Europe / Radio Liberty's Kyrgyz service, \href{http://www.azattyk.org}{azattyk.org}.  The RFERL corpus was built using a script that scraped the archives for all articles from 2010, which include articles on a wide variety of topics, from sports to politics and from culture to current events.

@@

\begin{table*}[htbp]
	\centering
	\begin{tabular}{lrrrr}
		\toprule
		Corpus           & Tokens    & Known       & Naïve coverage (\%) & Mean ambiguity\\
		\midrule 
		Kyrgyz Wikipedia & 329,524   & 270,668     & 82.1 $\pm$ 3.2      & 2.35\\
		RFERL Kyrgyzstan & 4,112,558 & 3,614,193   & 87.9 $\pm$ 1.2      & 2.43\\
		\bottomrule
	\end{tabular}
	\caption{Naïve coverage and mean ambiguity of the analyser over two test corpora.}
	\label{table:cov}
\end{table*}

@@

%FIXME: * Some remaining transducer issues
% maybe fixed now?
Also, despite the good coverage, there are still a number of grammatical forms that have not been implemented into the transducer.  For example, one such form is historically the negative of the eyewitness past tense, but is now used (mostly with a question clitic) in a certain type of modal expression;\footnote{A surprised contradictory expression, like [Идиштерди жуубадымбы ?] \eng{Did I not just wash the dishes ??}} this cannot be tagged as the negative of the eyewitness past tense, since a suppletive form is tagged that way.  Other shortcomings of the transducer are that it is not set up to deal with affixes on abbreviations or numerals/digits (e.g., [АКШнын] \eng{The USA's}, [100гө] \eng{to 100}) or capitalisation changes on lemmas that occur in different word classes (e.g., [Финландия] \eng{Finland}, but [финландиялык] \eng{Finnish}).

%FIXME: e.g.?
%these issues is fixed!
%Also, despite the good coverage, there are still a number of grammatical forms that have not been implemented into the transducer.  One of these is the irregular combination of 3rd person non-past finite verb plus the yes/no question particle.  For example, the third person non-past of /бар/ \eng{go} is [барат], but when the question particle -/BI/ is added, the resulting form is [барабы] instead of *[баратпы] (the transducer's current output).  There are also currently some problems handling the progressive aspect auxiliary /жат/ when it grammaticalises as a suffix with a couple verbs (/бар/ and /кел/) which often use a different participial form with this auxiliary than most verbs do.

>>>>

@@

The understanding of Kyrgyz grammar employed to construct the transducer for this project was gained from the Kyrgyz knowledge of two of our authors.  Mirlan Ipasov is a native speaker of Kyrgyz, and Jonathan Washington is a theoretical and descriptive linguist fluent in Kyrgyz and knowledgeable about Turkic languages.  Some grammar sources were consulted, such as \cite{hebertpoppe1963}, \cite{usonalievomuraliev2003}, \cite{qudaybergenov1980}, \cite{somfaikara2003}, and \cite{imart1981}, but they were largely not relied on due to their approaches to Kyrgyz grammar.  Some dictionaries were also consulted, including \cite{jumakunova2005} and \citetwoyears{yudakhin1957}{yudakhin1965}.

\subsection{Morphological transducers}
The objective of a morphological transducer is twofold. Firstly to take surface forms (e.g., алдым) and generate all possible lexical forms, and secondly to take lexical forms (e.g.,  ал{\tt {\small <v><tv><ifi><p1><sg>}}, алд{\tt {\small <n><px1sg><nom>}}, etc.) and generate one or more surface forms.

@@

\section{Description}
The tagset consists of 127 separate tags, 19 covering the main parts of speech (noun, verb, adjective, adverb, postposition, etc.) and 108 covering morphological subcategorisation for e.g. case, number, person, possession, transitivity, tense-aspect-mood, etc. The tags are represented as multicharacter symbols, between less than `<' and greater than `>' symbols. The tagset is quite extensive and still not entirely stabilised, as such a full listing is not included here. However, the tags are listed in the source code of the transducer, along with comments describing their usage.
%FIXME: URL for source code?

@@

%TODO: for a future version
%\subsection{Challenges:  Copula, how to represent different tenses

@@

The first exception does not apply to Kyrgyz, as unlike other Turkic languages (including Tatar, Chuvash, and Kazakh), although it has morphophonological processes that work across the `whitespace' boundary, these are not represented in the orthography.

@@

%Future TODO (was a FIXME): why copular suffixes?

@@

\footnotetext{P1--P6 refer to different sets of personal suffixes; the terminology and specific numbers are based off of \citet[29]{hebertpoppe1963}.}

@@

\begin{table}
	\caption{Breakdown of number of stems in the lexicon by major part of speech. Minor categories, such as particle, modal and copula are left out.}\label{table:stems}
	\centering
	\begin{tabular}{lr}
		\toprule
		Part of speech & Number of stems\\
		\midrule
		Noun & 4,972\\
		Verb & 1,231 \\
		Adjective & 944\\
		Proper noun & 796\\
		Adverb & 295 \\
		Numeral & 63 \\
		Conjunction & 58 \\
		Postposition & 51 \\
		Pronoun & 29\\
		Determiner & 27\\
		\midrule
		Total: & 8,466 \\
		\bottomrule
	\end{tabular}
\end{table}

@@

To calculate the naïve coverage\footnote{Naïve coverage refers to the percentage of surface forms in a given corpora that receive at least one analysis.  Forms counted by this measure may have other analyses which are not delivered by the transducer.} of the analyser, two corpora were used.  The first corpus was a database dump of the Kyrgyz Wikipedia, dated 2011-09-23,\footnote{The exact name of the dump was \texttt{kywiki-20110923-pages-articles.xml}.} which was processed with the programs \texttt{aq-wikicrp} to extract sentences. The second was a corpus generated from the archives of Radio Free Europe / Radio Liberty's Kyrgyz service, \href{http://www.azattyk.org}{azattyk.org}.  The RFERL corpus was built using a script that scraped the archives for all articles from 2010, which include articles on a wide variety of topics, from sports to politics and from culture to current events.

@@

\begin{table*}[htbp]
	\centering
	\caption{Naïve coverage and mean ambiguity of the analyser over two test corpora.}
	\begin{tabular}{lrrrr}
		\toprule
		Corpus           & Tokens    & Known       & Naïve coverage (\%) & Mean ambiguity\\
		\midrule 
		Kyrgyz Wikipedia & 329,524   & 270,668     & 82.1 $\pm$ 3.2      & 2.35\\
		RFERL Kyrgyzstan & 4,112,558 & 3,614,193   & 87.9 $\pm$ 1.2      & 2.43\\
		\bottomrule
	\end{tabular}
	\label{table:cov}
\end{table*}

@@

Also, despite the good coverage, there are still a number of grammatical forms that have not been implemented into the transducer.  For example, one such form is historically the negative of the eyewitness past tense, but is now used (mostly with a question clitic) in a certain type of modal expression;\footnote{A surprised contradictory expression, like [Идиштерди жуубадымбы ?] \eng{Did I not just wash the dishes ??}} this cannot be tagged as the negative of the eyewitness past tense, since a suppletive form is tagged that way.  Other shortcomings of the transducer are that it is not set up to deal with affixes on abbreviations or numerals/digits (e.g., [АКШнын] \eng{The USA's}, [100гө] \eng{to 100}) or capitalisation changes on lemmas that occur in different word classes (e.g., [Финландия] \eng{Finland}, but [финландиялык] \eng{Finnish}).
